Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2018-09-12 RNARedPrint.tex
Make consistent again, edits, ...

\newcommand{\substitute}[2]{#1\!\oplus\!#2}
%\newcommand{\evalfor}[2]{#1\llbracket{}#2\rrbracket{}}
\newcommand{\evalfor}[2]{#1(#2)}

@@

\begin{abstractbox}
\begin{abstract}
  \parttitle{Background}
  The design of multi-stable RNA molecules has important applications in biology, medicine, and biotechnology. Synthetic design approaches profit strongly from effective in-silico methods, which can tremendously impact their cost and feasibility. \\
  \parttitle{Results} We revisit a central ingredient of most in-silico
  design methods: the sampling of sequences for the design of
  multi-target structures, possibly including pseudoknots. For this
  task, we present the efficient, tree decomposition-based algorithm
  \ourprog{}. Our fixed parameter tractable approach is underpinned by
  establishing the $\#${\sf P}-hardness of uniform sampling. Modeling
  the problem as a constraint network, \ourprog{} supports generic
  Boltzmann-weighted sampling for arbitrary additive RNA energy
  models; this enables the generation of RNA sequences meeting
  specific goals like expected free energies or \GCb-content. Finally,
  we empirically study general properties of the approach and generate
  biologically relevant multi-target Boltzmann-weighted designs for a
  common design benchmark, including pseudoknots. Generating seed sequences with \ourprog{}, we demonstrate significant improvements over the previously best multi-target sampling strategy.\\

@@

To answer these questions, we introduce a generic framework (illustrated in~Fig.~\ref{fig:workflow}) enabling efficient Boltzmann-weighted sampling over RNA sequences with multiple target structures. As main algorithmic contribution of this work, we devise a dynamic programming algorithm, based on a tree decomposition, to compute partition functions and sample sequences from the Boltzmann distribution%
% (Subsection~\ref{sec:PF})
. We show that these algorithms are \Def{fixed-parameter tractable (FPT)} for the \Def{treewidth},
in other words: they have polynomial time and space complexity for a fixed treewidth%
% Subsection~\ref{sec:complexity}
%
. Due to the generality of our method, we can moreover strongly limit this parameter in practice by using state-of-the-art tree decomposition algorithms.
%Uniform sampling is handled as a special case of Boltzmann sampling, where each valid sequence receives energy zero and---consequently---computing partition functions specializes to counting.
By evaluating (partial) sequences in a weighted constraint
network, we support arbitrary multi-ary constraints and thus
arbitrarily complex energy models,
notably subsuming all commonly
used RNA energy models%
%(Subsection~\ref{sec:energy_models})
.  Moreover, we describe an \Def{adaptive
  sampling} strategy to control the free energies of the individual
target structures and \GCb\%%
% (SubSection~\ref{sec:multiBoltzmann})
. %

@@

\paragraph{Representing \GCb\%-content and energies as features.}
Formally, we define a \Def{feature $F$} as function on sequences, whose value is obtained by summing over an associated set of \Def{atomic feature contributions} $f_1, f_2\ldots$, briefly called \emph{contributions}, that each are associated with a restricted set of positions, such that
$$
F(S) =\!\!\!\!\sum_{\substack{\text{contribution $f_i$ of $F$}\\ \text{on positions $x_1,\ldots,x_p$}}}\!\!\!\! f_i(\{x_1\mapsto S_{x_1},\ldots,x_p\mapsto S_{x_p}\}).
$$

As simple example, the $\GCb\%$ feature is expressed by $n$ contributions $f^{\GC}_i$ where $f^{\GC}_i(\{i\mapsto S_i\})$ equals $-1$ if $S_i=\Gb$ or $\Cb$, and $0$ otherwise.

For the definition of energy features, we abstract an \Def{RNA (secondary) structure $R$ of  length $n$} as a set of \Def{base pairs} $(i,j)$, where
$1\leq i<j\leq n$ and for all different $(i,j), (i',j')\in R$: $\{i,j\}\cap\{i',j'\}=\emptyset$ (``degree $\leq$ 1'')---without precluding pseudoknots.
% We call an RNA structure $R$ \Def{non-crossing}, iff it does not
% contain any two different base pairs $(i,j)$ and $(i',j')$ such that
% $i\leq i'\leq j \leq j'.$
%
We call $S$ \Def{valid} for $R$, iff $\{S_i,S_j\}\in \B$ for
all $(i,j)\in R$, where $\B:=\left\{\{\Ab,\Ub\},\{\Gb,\Cb\},\{\Gb,\Ub\}\right\}$ describes the valid base pairs.

@@

@@

 \paragraph{Further preliminaries for complexity and algorithmic considerations} 
The complexity of the central problem depends fundamentally on the dependency structure that is imposed by the entire set of contributions.

@@

More generally, we consider dependencies induced by the complete set of contributions $\F := \bigcup_{F}\{f\mid f\text{ contribution of } F\}$.
Recall that each contribution, even if `formally' evaluated for entire sequences, depends only on a subset of all sequence positions, which we call the \Def{dependencies of $f$} and write as $\dep(f)$;
thus, $\dep(f)$ is the minimal subset of $\{1,\dots,n\}$ s.t.~$\forall S,S'\in\Sigma^n: f(S)=f(S') \text{ only if } (\forall i\in\dep(f): S_i=S_i')$.

Then, the set $\F$ of contributions induces a
  \Def{dependency graph} $G_\F$ on sequence positions by interpreting the dependencies as hyperedges, namely the hypergraph 
  $G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\})$. The \Def{tree decomposition} of $G_\F$ is fundamental for our algorithms.

@@

Our algorithms iteratively merge partial solutions. To capture this we introduce 
\Def{Partial RNA sequences} as words $\val$ in
$(\Sigma\cup\{?\})^n$, where the symbol '?' indicate unassigned positions; consequently, positions such that  $\val_i\in\Sigma$,
are called \Def{determined} by $\val$ ($1\leq i\leq n$).
%
We remark that (partial) sequences correspond to the more general concept of (partial) assignments in
constraint networks.

Since each contribution $f\in\F$
depends on only the subset $\dep(f)$ of sequence positions, one can
evaluate them for partial sequences $\val$ that determine (at least)
the nucleotides at all positions in $\dep(f)$. Thus for any such function $f$
and partial sequence $\val$, we know that $f(S)$ has the same value for any sequence $S\in \Sigma^n$ that equals $\val$ at its determined positions; thus, we define $\evalfor{f}{\val}$ as this common value.

@@

For the algorithms, it is convenient to encode the weight $\pi$ of each feature in its contributions, such that all contributions can be handled uniformly. For this purpose, we define the \Def{transformed contribution} 
$\weighttransform{f}{\pi}(S) := \ln(\pi)f(S),$
such that $exp(-\weighttransform{f}{\pi}(S))=\pi^{-f(S)}.$ 
%

@@

By these definitions, the network encodes the partition function $\partfun{\pi_0,\dots,\pi_k}$ of Eq.~(\ref{eq:mainproblem}); also confer Eq.~(\ref{eq:feature-energy-transformation}). Subsequently, we show how to efficiently compute this partition function based on the network.

\subsubsection*{Partition function and  stochastic backtrack algorithms}\label{sec:PF}

The partition function
over an RNA design network can be computed by dynamic programming based
on a tree decomposition of the network's dependency graph,
i.e.~essentially cluster tree elimination (CTE) \citep{Dechter2013}. Note that analogous algorithms could be constructed to count valid sequences or minimize energy.
%This yields fixed parameter tractable (FPT) algorithms, based on the treewidth.

@@

Our algorithms are formulated to process a \Def{cluster tree} of the
RNA design network, which is a tuple $(T,\chi,\phi)$, where
$(T,\chi)$ is a tree decomposition of $G_\F$, and $\phi(v)$ represents
a set of functions $f$, each uniquely assigned to a node $v\in T$;
$\dep(f)\subseteq\chi(v)$ and $\phi(v)\cap \phi(v')=\varnothing$ for
all $v\neq v'$.

@@

Our algorithms need to iterate over sets of partial sequences that determine specific sequence positions
from a set $\Y$; we denote the set of all partial sequences that exactly determine the positions in $\Y$ by $\partseqs(\Y)$.
Furthermore, given partial sequences $\val$ and
$\val'$, we define the \Def{combined partial sequence $\substitute{\val}{\val'}$} such that
$$
(\substitute{\val}{\val'})_i :=
\begin{cases}
  \val'_i & \text{if } \val'_i\in \Sigma\\
  \val_i & \text{otherwise}
\end{cases}
$$

\paragraph{FPT Computation of the partition function}
Given is the RNA design network $\network=(\X,\F)$ and its cluster
tree $(T,\chi,\phi)$.  
%
We assume several properties of this cluster tree, which otherwise could be easily obtained:
\begin{itemize}
\item
$T$ is
connected and contains a dedicated node $r$, with
$\chi(r)=\varnothing$ and $\phi(r)=\varnothing$. If such a root does not exist, it can be added to the tree decomposition and connected to one node in each connected component of $T$.
\item
all edges in the tree decompostion are oriented towards this root.
\item
all sets $\difference{u}{v}$ are singleton: for any given
cluster tree, an equivalent (in term of treewidth) cluster tree can
always be obtained by inserting at most $\Theta(|\X|)$ additional
clusters.
\end{itemize}

Algorithm~\ref{alg:pf} computes the partition function by passing
messages along the directed edges $u\to v$ (which point from 
child $u$ to its parent $v$). Each message $m$ is a function that depends
on the positions $\dep(m)\subseteq \X$ and yields a partition function
in $\real\cup\{\infty\}$. The message from $u$ to $v$ represents the
partition functions of the subtree of $u$ for all possible partial
sequences in $\partseqs(\separator{u}{v})$. Induction over $T$ lets us show
the correctness of the algorithm
(Supp. Mat.~\ref{appsec:correctness}).  After running
Alg.~\ref{alg:pf}, multiplying the 0-ary messages sent to the root $r$
yields the total partition function of the network (i.e.~due to proper encoding the partition function of our design problem)
\begin{math}
  \prod_{(u\to{}r)\in T} \evalfor{\Message{u}{r}}{\varnothing}.
\end{math}

@@

The partition functions can then direct a stochastic backtracking procedure to sample sequences from the Boltzmann distribution based on the given design network $\network$.
%
For an expanded cluster tree, after the messages $\Message{u}{v}$ for
the edges in the tree decomposition are generated by Algorithm \ref{alg:pf},
one can repeatedly call Algorithm~\ref{alg:sampling}, each time
randomly drawing another sequence from the Boltzmann distribution.

@@

\textbf{Move the approximation of more complex models from "Results" to this subsection}

%%!!!!!!!!!!!!!!! consider to merge into disciussion
%Going beyond our encoding of the base pair model energy features and the \GCb\% feature, the contributions provide more generality and allow expressing complex features of the
%sequences alone, e.g.\ rewarding or penalizing specific sequence
%motifs, as well as features depending on the target structures.
%Furthermore, constraints, which enforce or forbid features, are
%naturally expressed by assigning infinite penalties to invalid
%(partial) sequences (even if in implementations, explicit handling of constraints is advantageous). In particular, the %framework
%aptures various RNA energy models, they can be handled efficiently (in the sense of FPT), as long as the dependencies of their additive energy contributions are bounded.

@@

Complex \Def{loop-based}
 energy models---e.g.~the Turner model,
which among others includes energy terms for special loops and dangling
ends---can be encoded as straightforward extension of our outline of the base pair energy model.

@@

The arity of the introduced functions provides an important bound on the
treewidth of the network (and therefore computational
complexity). Thus, it is noteworthy that the base pair energy model
requires only binary functions; the stacking model, only quarternary
dependencies. This arity is increased in a few cases by the commonly
used Turner 2004 model~\citep{Turner2009} for encoding tabulated
special hairpin and interior loop contributions, which depend on up to
nine bases for the interior loops with a total of 5 unpaired bases
(``2x3'' interior loops)---all other energy contributions (like
dangling ends) still depend on at most four bases of the sequence.

@@

%Such a distribution can be induced by a simple modification of the functions described in Sec.~\ref{sec:energy_models}, where any energy function $E(\val)$ for a structure $\ell$ is replaced by $E'(\val):= \ln(\pi_\ell)\, E(\val)/\beta.$ 
Due to the construction of $\F$ of our RNA design network,
the probability of a sequence $S$ is thus proportional to
$\prod_{\ell=0}^{k} \pi_i^{-F_i(S)} = \prod_{f\in\F} f(S).$

@@

To capture the realistic Turner model $\EnergyTurner$ while keeping
a reasonable computational cost, we exploit a tight correlation between
$\EnergyTurner$ and the fitted stacking model $\EnergyStacking$
(Supp. Fig~\ref{fig:training-cor}). More precisely, we observed a
structure-specific affine dependency between the Turner and stacking
energy models, so that
$\EnergyTurner(S;R) \approx \gamma\cdot \EnergyStacking(S;R) + \delta$
for any structure $R$ and sequence $S$. We inferred the
$(\gamma,\delta)$ parameters from a set of sequences generated with
homogenous weights $w=e^{\beta}$, tuning only \GCb\% to a
predetermined value.  Finally, we adjusted the targeted energies
within our stacking model to
$\EnergyStacking^{\star} = (\EnergyTurner^{\star}- \delta)/\gamma$ in
order to reach, on average, the targeted energy
$\EnergyTurner^{\star}$ in the Turner model.

Fig.~\ref{fig:energydist-pk} illustrates the described strategy. For
the two target structures of Fig.~\ref{fig:energydist-pk}B,
Fig.~\ref{fig:energydist-pk}A shows the good fit between realistic
energies in the full-fledged Dirks and Pierce energy model for
pseudoknots (D\&P model) and energies in the stacking energy
model. For the shown fits we sampled $n=1\,000$ sequences, targeting a
\GCb\% of $60\%$.
%
For an example instance of the \texttt{Modena} benchmark with two
pseudoknotted target structures, Fig.~\ref{fig:energydist-pk}B shows
the Turner energy distributions of the single structures as they
result from sampling with different weight parameters. The figure
illustrates how our multidimensionnal Boltzmann sampling strategy can,
to a large extent, independently shift the Turner energies of sampled
sequences towards prescribed targets. See Supp. Mat.,
Fig.~\ref{fig:energydist} for a further example with three
pseudoknot-free target structures.

@@

\section*{Conclusion}
Motivated by the---here established---hardness of the problem, we
introduced a general framework, and efficient algorithms to design RNA
sequences that may adopt multiple target structures with fine-tuned
properties. Our method combines an FPT stochastic sampling algorithm
with multi-dimensional Boltzmann sampling over distributions
controlled by expressive RNA energy models.  Compared to the
previously best available sampling method (uniform sampling), this
approach generated significantly better seed sequences for instances
of an extensive multi-target design benchmark, including pseudoknots.

@@

\section*{Author's contributions}
  YP and SW initiated the project and worked out the algorithms.
  % SW---I guess we could be less detailed...
  YP contributed the multi-dimensional Boltzmann sampling background and the \#P-hardness proof.
  SW contributed the integration with the constraint network setting and its specialization to RNA design.
  YP and WW wrote the core software, while SW and SH contributed additional scripts. SH devised and conducted the computational experiments. All authors wrote, read, and approved the manuscript.
    % Text for this section \ldots
    % https://www.biomedcentral.com/getpublished/editorial-policies#authorship

>>>>

\newcommand{\substitute}[2]{#1\cup#2}
%\newcommand{\evalfor}[2]{#1\llbracket{}#2\rrbracket{}}
\newcommand{\evalfor}[2]{#1(#2)}

@@

\begin{abstractbox}
\begin{abstract}
  \parttitle{Background}
  The design of multi-stable RNA molecules has important applications in biology, medicine, and biotechnology. Synthetic design approaches profit strongly from effective in-silico methods, which can tremendously impact their cost and feasibility. \\
  \parttitle{Results} We revisit a central ingredient of most in-silico
  design methods: the sampling of sequences for the design of
  multi-target structures, possibly including pseudoknots. For this
  task, we present the efficient, tree decomposition-based algorithm
  \ourprog{}. Our fixed parameter tractable approach is underpinned by
  establishing the $\#${\sf P}-hardness of uniform sampling. Modeling
  the problem by specializing constraint networks, \ourprog{} supports generic
  Boltzmann-weighted sampling for arbitrary additive RNA energy
  models; this enables the generation of RNA sequences meeting
  specific goals like expected free energies or \GCb-content. Finally,
  we empirically study general properties of the approach and generate
  biologically relevant multi-target Boltzmann-weighted designs for a
  common design benchmark, including pseudoknots. Generating seed sequences with \ourprog{}, we demonstrate significant improvements over the previously best multi-target sampling strategy.\\

@@

To answer these questions, we introduce a generic framework (illustrated in~Fig.~\ref{fig:workflow}) enabling efficient Boltzmann-weighted sampling over RNA sequences with multiple target structures. As main algorithmic contribution of this work, we devise a dynamic programming algorithm, based on a tree decomposition, to compute partition functions and sample sequences from the Boltzmann distribution%
% (Subsection~\ref{sec:PF})
. We show that these algorithms are \Def{fixed-parameter tractable (FPT)} for the \Def{treewidth},
in other words: they have polynomial time and space complexity for a fixed treewidth%
% Subsection~\ref{sec:complexity}
%
. Due to the generality of our method, we can moreover strongly limit this parameter in practice by using state-of-the-art tree decomposition algorithms.
%Uniform sampling is handled as a special case of Boltzmann sampling, where each valid sequence receives energy zero and---consequently---computing partition functions specializes to counting.
By evaluating (partial) sequences in a specialized weighted constraint
network, we support arbitrary multi-ary constraints and thus
arbitrarily complex energy models,
notably subsuming all commonly
used RNA energy models%
%(Subsection~\ref{sec:energy_models})
.  Moreover, we describe an \Def{adaptive
  sampling} strategy to control the free energies of the individual
target structures and \GCb\%%
% (SubSection~\ref{sec:multiBoltzmann})
. %

@@

\paragraph{Representing \GCb\%-content and energies as features.}
Formally, we define a \Def{feature $F$} as function on sequences, whose value is obtained by summing over an associated set of \Def{atomic feature contributions} $f_1, f_2\ldots$, briefly called \emph{contributions}, that each are associated with a restricted set of positions, its \emph{dependencies $\dep(f)$}, such that
$$
F(S) =\!\!\!\!\!\!\sum_{\substack{\text{contribution $f_i$ of $F$}\\\text{with $\dep(f)=\{x_1,\ldots,x_p\}$}}}\!\!\!\!\!\! f_i(\{x_1\mapsto S_{x_1},\ldots,x_p\mapsto S_{x_p}\}).
$$

As simple example, the $\GCb\%$ feature is expressed by $n$ contributions $f^{\GCb}_i$ where $f^{\GCb}_i(\{i\mapsto S_i\})$ equals $-1$ if $S_i=\Gb$ or $\Cb$, and $0$ otherwise.

For the definition of energy features, we abstract an \Def{RNA (secondary) structure $R$ of  length $n$} as a set of \Def{base pairs} $(i,j)$, where
$1\leq i<j\leq n$ and for all different $(i,j), (i',j')\in R$: $\{i,j\}\cap\{i',j'\}=\emptyset$ (``degree $\leq$ 1'')---without precluding pseudoknots.
% We call an RNA structure $R$ \Def{non-crossing}, iff it does not
% contain any two different base pairs $(i,j)$ and $(i',j')$ such that
% $i\leq i'\leq j \leq j'.$

@@

Typically, such RNA energy models consider a sequence $S$ \Def{valid} for $R$, iff $\{S_i,S_j\}\in \B$ for
all $(i,j)\in R$, where $\B:=\left\{\{\Ab,\Ub\},\{\Gb,\Cb\},\{\Gb,\Ub\}\right\}$ describes the valid base pairs.

@@

\paragraph{Dependency structure and complexity}
%The complexity of the central problem depends fundamentally on the dependency structure that is imposed by the entire set of contributions.

@@

More generally, we consider dependencies induced by the complete set of contributions $\F := \bigcup_{\ell}\{f\mid f\text{ contribution of } F_\ell\}$.
%
%Let us denote the \emph{dependencies of a contribution $f$} as \dep{f}.
%
The set $\F$ of contributions induces a
  \Def{dependency graph} $G_\F$ on sequence positions by interpreting the dependencies as hyperedges, namely the hypergraph 
  $G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\})$. The \Def{tree decomposition} of $G_\F$ is fundamental for our algorithms.

@@

To formalize our algorithms, which iteratively merges evaluations of partial solutions, we extend the idea of atomic feature contributions, which are evaluated at sets of the form $\{x_1\mapsto v_1, \dots, x_p\mapsto v_d\}$. Let us call the latter object \Def{partial sequence}; they are handy to specify partial knowledge on the entire sequence at some point of the algorithm. Easily, we can extend the definition of contributions $f$ to sets $\{x_1\mapsto v_1, \dots, x_p\mapsto v_d\}$, where $\{x_1\dots x_d\}$ is any super-set of $\dep(f)$ by ignoring the superfluous assignments $x\mapsto v$, where $x\not\in\dep(f)$.

Our algorithms specialize the idea of cluster tree elimination (CTE)~\citep{Dechter2013}, which operates on constraint networks. In this correspondence, (partial) sequences specialize (partial) assignments and the constraint network would be given by variables for each sequence position, constraints due to valid base pairing, and the set of atomic feature contributions $\cal F$.

@@

For algorithmic treatment, it is convenient to encode the weight $\pi$ of each feature in its contributions, such that all contributions can be handled uniformly. For this purpose, we define the \Def{transformed contribution} 
$\weighttransform{f}{\pi}(S) := \ln(\pi)f(S),$
such that $exp(-\weighttransform{f}{\pi}(S))=\pi^{-f(S)}.$ 
%

@@

By these definitions, the set $\F$ encodes the partition function $\partfun{\pi_0,\dots,\pi_k}$ of Eq.~(\ref{eq:mainproblem}); also confer Eq.~(\ref{eq:feature-energy-transformation}). 
%Subsequently, we show how to efficiently compute this partition function based on the 
%corresponding dependency graph and its tree decomposition.

\subsubsection*{Partition function and stochastic backtracking}\label{sec:PF}

We compute the partition function (as specified by $\F$) by dynamic programming based
on a tree decomposition of the $\F$'s dependency graph.
Note that analogous algorithms could be constructed to count valid sequences or minimize energy.
%This yields fixed parameter tractable (FPT) algorithms, based on the treewidth.

@@

Our algorithms are formulated to process a \Def{cluster tree}, which is a tuple $(T,\chi,\phi)$, where
$(T,\chi)$ is a tree decomposition of $G_\F$, and $\phi(v)$ represents
a set of functions $f$, each uniquely assigned to a node $v\in T$;
$\dep(f)\subseteq\chi(v)$ and $\phi(v)\cap \phi(v')=\varnothing$ for
all $v\neq v'$.

@@

Since our algorithms iterate over specific sets of sequence positions, we moreover define
the \Def{set of partial sequences $\partseqs(\Y)$ determining $\Y\subseteq\{1,\dots,n\}$} as
$$
\partseqs(\Y)=\{ x\mapsto v \mid \text{$x\in\Y$ and $v\in \{\Ab,\Cb,\Gb,\Ub\}$}\}.
$$
%
We assume several properties of the given cluster tree (reflecting $\F$), which otherwise could be easily obtained:
\begin{itemize}
\item
$T$ is
connected and contains a dedicated node $r$, with
$\chi(r)=\varnothing$ and $\phi(r)=\varnothing$. If such a root does not exist, it can be added to the tree decomposition and connected to one node in each connected component of $T$.
\item
all edges in the tree decompostion are oriented towards this root.
\item
all sets $\difference{u}{v}$ are singleton: for any given
cluster tree, an equivalent (in term of treewidth) cluster tree can
always be obtained by inserting at most $\Theta(|\X|)$ additional
clusters.
\end{itemize}

Algorithm~\ref{alg:pf} computes the partition function by passing
messages along the directed edges $u\to v$ (which point from 
child $u$ to its parent $v$). Each message $m$ is a function that depends
on the positions $\dep(m)\subseteq \X$ and yields a partition function
in $\real$. The message from $u$ to $v$ represents the
partition functions of the subtree of $u$ for all possible partial
sequences in $\partseqs(\separator{u}{v})$. Induction over $T$ lets us show
the correctness of the algorithm
(Supp. Mat.~\ref{appsec:correctness}).  After running
Alg.~\ref{alg:pf}, multiplying the 0-ary messages sent to the root $r$
yields the total partition function (i.e.~due to proper encoding the partition function of our design problem)
\begin{math}
  \prod_{(u\to{}r)\in T} \evalfor{\Message{u}{r}}{\varnothing}.
\end{math}

@@

The partition functions can then direct a stochastic backtracking procedure to sample sequences from the Boltzmann distribution (according to $\F$).
%
For an expanded cluster tree, after the messages $\Message{u}{v}$ for
the edges in the tree decomposition are generated by Algorithm \ref{alg:pf},
one can repeatedly call Algorithm~\ref{alg:sampling}, each time
randomly drawing another sequence from the Boltzmann distribution.

@@

@@

So far, we discussed sampling only in base pair energy models, although one is finally interested in targeting energies in realistic energy models like the Turner model or pseudoknot models like {\tt HotKnots}~\citep{Ren2005}.
Notably, for this purpose, the discussed sampling strategy can be extended in two ways: 1) one can directly sample based on more expressive energy models, 2) sampling in a simple energy model can be used to approximate sampling in more complex models.

\paragraph{Sampling based on more expressive energy features}
Complex \Def{loop-based}
 energy models---e.g.~the Turner model,
which among others includes energy terms for special loops and dangling
ends---can be encoded as straightforward extension of our outline of the base pair energy model.

@@

The arity of the introduced functions bounds the
treewidth of the dependency graph. %(and therefore computational complexity).
Thus, it is noteworthy that the base pair energy model
requires only binary functions; the stacking model, only quarternary
dependencies. This arity is increased in a few cases by the commonly
used Turner 2004 model~\citep{Turner2009} for encoding tabulated
special hairpin and interior loop contributions, which depend on up to
nine bases for the interior loops with a total of 5 unpaired bases
(``2x3'' interior loops)---all other energy contributions (like
dangling ends) still depend on at most four bases of the sequence.

\paragraph{Approximating Turner Energy using Simpler Energy Models}
To capture the realistic Turner model $\EnergyTurner$ while keeping
a reasonable computational cost, we exploit a tight correlation between
$\EnergyTurner$ and the fitted stacking model $\EnergyStacking$
(Supp. Fig~\ref{fig:training-cor}). More precisely, we observed a
structure-specific affine dependency between the Turner and stacking
energy models, so that
$\EnergyTurner(S;R) \approx \gamma\cdot \EnergyStacking(S;R) + \delta$
for any structure $R$ and sequence $S$. We inferred the
$(\gamma,\delta)$ parameters from a set of sequences generated with
homogenous weights $w=e^{\beta}$, tuning only \GCb\% to a
predetermined value.  Finally, we adjusted the targeted energies
within our stacking model to
$\EnergyStacking^{\star} = (\EnergyTurner^{\star}- \delta)/\gamma$ in
order to reach, on average, the targeted energy
$\EnergyTurner^{\star}$ in the Turner model.

@@

%Such a distribution can be induced by a simple modification of the functions described in Sec.~\ref{sec:energy_models}, where any energy function $E(\val)$ for a structure $\ell$ is replaced by $E'(\val):= \ln(\pi_\ell)\, E(\val)/\beta.$ 
Due to the construction of $\F$,
the probability of a sequence $S$ is thus proportional to
$\prod_{\ell=0}^{k} \pi_i^{-F_i(S)} = \prod_{f\in\F} f(S).$

@@

Fig.~\ref{fig:energydist-pk} illustrates how well complex realistic energy models can be approximated based on simpler, but better tractable ones. For
the two target structures of Fig.~\ref{fig:energydist-pk}B,
Fig.~\ref{fig:energydist-pk}A shows the good fit between realistic
energies in the full-fledged Dirks and Pierce energy model for
pseudoknots (D\&P model) and energies in the stacking energy
model. For the shown fits we sampled $n=1\,000$ sequences, targeting a
\GCb\% of $60\%$.
%
For an example instance of the \texttt{Modena} benchmark with two
pseudoknotted target structures, Fig.~\ref{fig:energydist-pk}B shows
the Turner energy distributions of the single structures as they
result from sampling with different weight parameters. The figure
illustrates how our multidimensionnal Boltzmann sampling strategy can,
to a large extent, independently shift the Turner energies of sampled
sequences towards prescribed targets. See Supp. Mat.,
Fig.~\ref{fig:energydist} for a further example with three
pseudoknot-free target structures.

@@

\section*{Conclusion}
%%!!!!!!!!!!!!!!! consider to merge into disciussion
%Going beyond our encoding of the base pair model energy features and the \GCb\% feature, the contributions provide more generality and allow expressing complex features of the
%sequences alone, e.g.\ rewarding or penalizing specific sequence
%motifs, as well as features depending on the target structures.
%Furthermore, constraints, which enforce or forbid features, are
%naturally expressed by assigning infinite penalties to invalid
%(partial) sequences (even if in implementations, explicit handling of constraints is advantageous). In particular, the %framework
%aptures various RNA energy models, they can be handled efficiently (in the sense of FPT), as long as the dependencies of their additive energy contributions are bounded.

Motivated by the---here established---hardness of the problem, we
introduced a general framework, and efficient algorithms to design RNA
sequences that may adopt multiple target structures with fine-tuned
properties. Our method combines an FPT stochastic sampling algorithm
with multi-dimensional Boltzmann sampling over distributions
controlled by expressive RNA energy models.  Compared to the
previously best available sampling method (uniform sampling), this
approach generated significantly better seed sequences for instances
of an extensive multi-target design benchmark, including pseudoknots.

@@

\section*{Author's contributions}
  YP and SW initiated the project and worked out the algorithms.
  % SW---I guess we don't need to be super-detailed, otherwise something like...
  %YP contributed the multi-dimensional Boltzmann sampling background and the \#P-hardness proof.
  %SW contributed the integration with the constraint network setting and its specialization to RNA design.
  YP and WW wrote the core software, while SW and SH contributed additional scripts. SH devised and conducted the computational experiments. All authors wrote, read, and approved the manuscript.
    % Text for this section \ldots
    % https://www.biomedcentral.com/getpublished/editorial-policies#authorship
